---
title: "AI Impact and Employment Growth: Mapping the Future of Work"
author: "Song Bai"
format: 
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
---

## Introduction

Artificial Intelligence (AI) is reshaping the modern workforce. This report explores how AI's projected impact on job tasks correlates with anticipated employment growth or decline over the next decade. Specifically, we ask:

> **Is there a relationship between AI impact and projected employment growth by occupation, and what does it reveal about the future of work?**

## Data Sources

```{r, message=FALSE}
# Load required libraries
if (!requireNamespace("httr", quietly = TRUE)) install.packages("httr")
library(httr)

# Define paths
dir_path <- "data/project"
zip_filename <- "from-data-entry-to-ceo-the-ai-job-threat-index.zip"
zip_path <- file.path(dir_path, zip_filename)

# Create directory if it doesn't exist
if (!dir.exists(dir_path)) {
  dir.create(dir_path, recursive = TRUE)
}

# Check if file already exists
if (!file.exists(zip_path)) {
  
  # Load Kaggle credentials from ~/.kaggle/kaggle.json
  kaggle_json_path <- file.path(Sys.getenv("HOME"), ".kaggle", "kaggle.json")
  
  if (!file.exists(kaggle_json_path)) {
    stop("Kaggle API key not found at ~/.kaggle/kaggle.json")
  }
  
  kaggle_creds <- jsonlite::fromJSON(kaggle_json_path)
  username <- kaggle_creds$username
  key <- kaggle_creds$key
  
  # Prepare request
  kaggle_url <- "https://www.kaggle.com/api/v1/datasets/download/manavgupta92/from-data-entry-to-ceo-the-ai-job-threat-index"
  
  message("Downloading dataset from Kaggle using R...")
  
  response <- httr::GET(
    url = kaggle_url,
    authenticate(username, key),
    httr::write_disk(zip_path, overwrite = TRUE),
    httr::progress()
  )
  
  if (httr::status_code(response) != 200) {
    stop("Failed to download dataset. Status code: ", httr::status_code(response))
  }
  
} else {
  message("Dataset already downloaded.")
}

# Unzip if not already unzipped
csv_files <- list.files(dir_path, pattern = "\\.csv$", full.names = TRUE)
if (length(csv_files) == 0 && file.exists(zip_path)) {
  message("Unzipping dataset...")
  unzip(zip_path, exdir = dir_path)
} else {
  message("Dataset already unzipped or no zip found.")
}

# read data into r
ai <- read.csv("data/project/My_Data.csv")
employment <- read.csv("data/project/Employment Projections.csv")

```

- **AI Impact Dataset**: Contains AI workload ratios and task information by job title.
- **Employment Projections Dataset**: Includes projected job growth, median wages, and education requirements by occupation.

Both datasets required cleaning and merging based on job titles, which often involved multiple aliases or fuzzy matches.

## Data Processing

```{r, message=FALSE, warning=FALSE}
library(dplyr)
library(stringr)
library(tidyr)

# Clean job titles
ai <- ai |>
  mutate(Job.title.clean = str_to_lower(str_trim(Job.titiles)))

employment <- employment |>
  mutate(Occupation.Title.clean = str_to_lower(str_trim(Occupation.Title)))

# Occupation.Title contains multiple titles separated by *, we split those:
employment <- employment |>
  separate_rows(Occupation.Title.clean, sep = "\\*") |>
  mutate(Occupation.Title.clean = str_trim(Occupation.Title.clean))

# Use a fuzzy join as exact matches may not work
# install.packages("fuzzyjoin")
library(fuzzyjoin)

combined <- stringdist_inner_join(ai, employment,
                                  by = c("Job.title.clean" = "Occupation.Title.clean"),
                                  method = "jw", max_dist = 0.15) 

# remove helper columns
combined <- combined |> select(-Job.title.clean, -Occupation.Title.clean)

```

## AI Impact vs. Projected Employment Growth 

To explore the relationship between AI exposure and job market trends, I created a scatter plot comparing the AI Impact (%) of each occupation with its projected employment growth (2023–2033). AI impact scores were converted from percentage strings to numeric values, and employment growth was taken directly from U.S. Bureau of Labor Statistics projections.

```{r, message=FALSE}
library(ggplot2)
library(readr)
# Convert AI Impact % to numeric
combined <- combined |>
  mutate(AI.Impact.Numeric = as.numeric(str_remove(AI.Impact, "%")),
         Employment.Growth = Employment.Percent.Change..2023.2033)

# Scatter plot: AI impact vs. employment growth
ggplot(combined, aes(x = AI.Impact.Numeric, y = Employment.Growth)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "lm", se = FALSE, color = "red") +
  labs(title = "AI Impact vs. Projected Employment Growth",
       x = "AI Impact (%)",
       y = "Employment Growth (2023–2033, %)",
       caption = "Data: AI Impact and U.S. Employment Projections") +
  theme_minimal()
```

The chart shows a wide dispersion of data points, with no strong visual trend indicating a direct relationship between AI exposure and job growth. The red regression line represents a linear model, which shows only a slight negative slope. This suggests that, on average, occupations with higher AI impact tend to have slightly lower projected growth, but the correlation is weak.

To quantitatively assess the relationship between AI exposure and projected job growth, I conducted a Pearson correlation test between:

- `combined$AI.Impact.Numeric` (AI impact score as a percentage), and

- `combined$Employment.Growth` (projected percent change in employment from 2023 to 2033).

```{r, results='hide'}
# Correlation coefficient
cor.test(combined$AI.Impact.Numeric, combined$Employment.Growth)
```

### Test Results

> t = -6.7359, df = 4124, p-value = 1.854e-11

> correlation coefficient = -0.104

> 95% CI: [-0.134, -0.074]

The correlation coefficient is -0.104, indicating a very weak negative correlation between AI impact and employment growth. The p-value < 0.0001 confirms this relationship is statistically significant, meaning the likelihood of observing this pattern by chance is extremely low. However, the magnitude of the correlation is small, suggesting that AI exposure explains very little of the variation in employment growth across occupations. 

These results reinforce the earlier visual findings. While AI may play a role in shaping job trajectories, it is not the sole or dominant factor influencing projected employment changes. Some highly impacted jobs are expected to grow — possibly due to AI augmentation rather than replacement. 

### Interactive Visualization I: AI Impact vs. Employment Growth by Domain

```{r, message=FALSE, warning=FALSE}
library(ggplot2)
library(plotly)

# Base ggplot
p <- ggplot(combined, aes(x = AI.Impact.Numeric, 
                          y = Employment.Growth,
                          color = Domain,
                          text = paste(
                            "Job Title:", Job.titiles,
                            "<br>AI Impact:", AI.Impact,
                            "<br>Employment Growth:", Employment.Growth, "%",
                            "<br>Domain:", Domain))) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(
    title = "AI Impact vs. Employment Growth (Interactive by Domain)",
    x = "AI Impact (%)",
    y = "Employment Growth (2023–2033, %)",
    color = "Occupational Domain"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

# Convert to interactive plot
ggplotly(p, tooltip = "text")

```

To enhance interpretability and user engagement, the scatter plot of AI impact versus projected employment growth was upgraded to an interactive graphic using the `plotly` library in R. In this version of the chart, each point represents an individual occupation. Points are color-coded by occupational domain, such as Healthcare, IT, or Clerical work. Users can hover over points to reveal job-specific details, including:

- Job title

- AI impact score (%)

- Projected employment growth (%)

- Occupational domain

This visualization helps bridge the gap between macro-level insights (e.g., AI is negatively correlated with growth) and micro-level stories (e.g., specific job roles that defy the trend). It supports a richer understanding of how AI may transform work differently across occupational domains.

### Interactive Visualization II: Faceted by Occupational Domain

```{r, message=FALSE, warning=FALSE}
# Load libraries
library(ggplot2)
library(plotly)

# Base ggplot
p <- ggplot(combined, aes(x = AI.Impact.Numeric, y = Employment.Growth)) +
  geom_point(aes(color = Domain, text = paste("Job:", Job.titiles,
                                              "<br>AI Impact:", AI.Impact,
                                              "<br>Growth:", round(Employment.Growth, 2), "%")), 
             alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  facet_wrap(~ Domain, scales = "free_y") +
  labs(
    title = "AI Impact vs. Employment Growth by Occupational Domain",
    x = "AI Impact (%)",
    y = "Employment Growth (2023–2033, %)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

# Convert to interactive plotly
ggplotly(p, tooltip = "text")

```

To explore the relationship between AI impact and projected employment growth across occupational categories, we created an interactive faceted plot using the `ggplot2` and `plotly` packages in R. This visualization enables users to engage dynamically with the data, filtering and examining patterns by domain and occupation.

In the plot, each point represents an occupation, colored by its associated domain (e.g., Data & IT, Healthcare, Administrative). The x-axis indicates the AI impact score (as a percentage), while the y-axis shows the projected employment growth between 2023 and 2033. A linear trend line is included in each facet to highlight overall directionality within each domain.

With this interactive plot, you can

- Use Hover tooltips to see the job title, AI impact percentage, and employment growth.

- Zoom for deeper exploration within specific domains.

This visualization reveals that some domains like construction show growth despite high AI impact, suggesting potential augmentation rather than replacement.

### Key Occupations

To highlight the occupational roles most likely to thrive in an AI-driven economy, we identified “key occupations” that meet two criteria:

- High AI Impact: ≥ 80% estimated task exposure to AI.

- Positive Employment Growth: ≥ 5% projected increase between 2023 and 2033.

These thresholds isolate occupations that are both heavily impacted by AI technologies and expected to expand in workforce demand, implying potential augmentation rather than replacement.

The interactive table below presents these key occupations, reveals a diverse set of occupations across multiple domains—particularly those leveraging data, communication, and technical skills. These roles demonstrate how AI is not solely a threat but also a driver of transformation and opportunity for certain career paths.

Use the interactive table below to sort, filter, and explore these roles in more detail.

```{r}
# Define thresholds
impact_threshold <- 80   # AI impact above 80%
growth_threshold <- 5   # Employment growth above 5%

# Flag key occupations
combined <- combined |>
  mutate(Key.Occupation = ifelse(AI.Impact.Numeric >= impact_threshold & 
                                   Employment.Growth >= growth_threshold, 
                                 TRUE, FALSE))

# Create the Table
key_jobs <- combined |>
  filter(Key.Occupation == TRUE) |>
  select(Job.titiles, Domain, AI.Impact.Numeric, Employment.Growth,
         Median.Annual.Wage.2024, Typical.Entry.Level.Education) |>
  arrange(desc(AI.Impact.Numeric), desc(Employment.Growth)) |>
  unique() 

# Interactive Table
DT::datatable(
  key_jobs,
  options = list(pageLength = 10),
  caption = "Key Occupations with High AI Impact and Employment Growth"
)
```

### Occupational Clustering

We applied k-means clustering to group jobs based on three key standardized variables:

- AI Impact (% of tasks potentially affected by AI),

- Projected Employment Growth (2023–2033), and

- Median Annual Wage (2024).

#### Methodology

After standardizing the variables and removing missing values, we used k-means clustering (k = 3) to identify latent occupational groups. These clusters were then interpreted and labeled based on their centroid characteristics:

- Transforming: High AI impact, moderate to high growth — roles likely to be augmented or significantly reshaped by AI.

- Declining: High AI impact, low or negative growth — roles at risk of automation-driven displacement.

- Safe: Lower AI impact and relatively stable growth — roles that appear resistant to AI-driven disruption.

#### Results

The scatterplot below visualizes these clusters. Each point represents an occupation, colored by its assigned cluster label. The distribution illustrates clear structural patterns. Many data-intensive and analytical roles are “Transforming” — affected by AI but growing in demand. Several clerical and administrative roles cluster in the “Declining” group, marked by high AI exposure and shrinking employment projections. “Safe” roles tend to be more interpersonal or physical in nature, with lower AI impact and stable demand. 

```{r, warning=FALSE, results='hide'}
# 1: Prepare Data for Clustering
# Clean wage to numeric
combined <- combined |>
  mutate(Wage = as.numeric(gsub(",", "", Median.Annual.Wage.2024)))

# Select and scale features
clustering_data <- combined |>
  select(AI.Impact.Numeric, Employment.Growth, Wage) |>
  scale()  # standardize

clustering_data_clean <- clustering_data |>
  as.data.frame() %>%
  filter(complete.cases(.))  # removes rows with NA
combined_clean <- combined[complete.cases(clustering_data), ]


# 2: Run k-means (3 clusters)
set.seed(123)
kmeans_result <- kmeans(clustering_data_clean, centers = 3, nstart = 25)

# Add cluster assignments to original data
combined_clean$Cluster <- as.factor(kmeans_result$cluster)

# 3: Label Clusters
aggregate(cbind(AI.Impact.Numeric, Employment.Growth, Wage) ~ Cluster, 
          data = combined_clean, 
          FUN = mean)
```

```{r}
combined_clean <- combined_clean |>
  mutate(Cluster.Label = recode(Cluster,
                                `1` = "Transforming",
                                `2` = "Declining",
                                `3` = "Safe"))

# 4: Visualize the Clusters
# Load necessary library
library(plotly)

# Create the ggplot object
p <- ggplot(combined_clean, aes(x = AI.Impact.Numeric, 
                                y = Employment.Growth, 
                                color = Cluster.Label,
                                text = paste("Job:", Job.titiles,
                                             "<br>AI Impact:", round(AI.Impact.Numeric, 1), "%",
                                             "<br>Growth:", round(Employment.Growth, 1), "%",
                                             "<br>Wage: $", Wage))) +
  geom_point(alpha = 0.7, size = 2) +
  labs(
    title = "Clustered Occupation Types",
    x = "AI Impact (%)",
    y = "Employment Growth (2023–2033, %)",
    color = "Occupation Type"
  ) +
  theme_minimal()

# Convert to interactive plot
ggplotly(p, tooltip = "text")

```

## Top 10 At-Risk and AI-Enhanced Occupations

```{r, warning=FALSE}
# 1: Create Ranking Tables
# Ensure numeric AI Impact and Growth
combined <- combined |>
  mutate(
    AI.Impact.Numeric = as.numeric(gsub("%", "", AI.Impact)),
    Employment.Growth = Employment.Percent.Change..2023.2033,
    Wage = as.numeric(gsub(",", "", Median.Annual.Wage.2024))
  )

# Top 10 AI-Enhanced
ai_enhanced <- combined |>
  filter(AI.Impact.Numeric >= 80, Employment.Growth > 5) |>
  arrange(desc(AI.Impact.Numeric), desc(Employment.Growth)) |>
  select(Job.titiles, Domain, AI.Impact.Numeric, Employment.Growth, Wage) |>
  head(10)

# Top 10 At-Risk
at_risk <- combined |>
  filter(AI.Impact.Numeric >= 80, Employment.Growth < 0) |>
  arrange(desc(AI.Impact.Numeric), Employment.Growth) |>
  select(Job.titiles, Domain, AI.Impact.Numeric, Employment.Growth, Wage) |>
  head(10)

# 2: Chart
ai_enhanced$Category <- "AI-Enhanced"
at_risk$Category <- "At-Risk"

top_jobs <- bind_rows(ai_enhanced, at_risk)

ggplot(top_jobs, aes(x = reorder(Job.titiles, AI.Impact.Numeric), 
                     y = AI.Impact.Numeric, 
                     fill = Category)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~ Category, scales = "free_y") +
  labs(
    title = "Top 10 At-Risk and AI-Enhanced Jobs",
    x = "Job Title",
    y = "AI Impact (%)"
  ) +
  theme_minimal()

```

We identified two sets of occupations: those with high AI impact (≥80%) and positive employment growth (>5%), and those with high AI impact but negative projected employment growth. These were categorized as AI-Enhanced and At-Risk jobs, respectively. The plot above presents a side-by-side comparison of the top 10 occupations in each category. Occupations are ranked by their AI impact, and the faceted view allows for clear distinction between the two trajectories. 

AI-Enhanced roles like Data Operators or Database Developers are expected to grow despite high exposure to AI, suggesting that AI might be complementing rather than replacing these positions. In contrast, At-Risk roles like Data Collectors or Data Clerks show declining employment prospects, highlighting the potential displacement effect of automation and AI technologies.

This comparison underscores the nuanced role of AI: while some jobs are being transformed and elevated, others are more vulnerable. These findings contribute to the broader understanding of how AI will likely reconfigure occupational demand, not uniformly eliminate or expand it.

